# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

1. Load the data
```{r}
data <- read.csv( "activity.csv", header = TRUE )
```

2. Reformat the date column as date data type
```{r}
data$date <- as.Date( data$date )
```

## What is mean total number of steps taken per day?

1. Create a new table containing total steps by date
```{r}
library( 'plyr' )
dailyTotal <- ddply( data, .(date), summarize, steps = sum( steps ) )
```

2. Plot a histogram of the total number of steps taken each day
```{r}
hist( dailyTotal$steps, breaks=12, main="Histogram of total number of steps taken per day", xlab="Steps per Day" )
```

3. Compute the mean and median total number of steps per day.
```{r}
meanSteps <- mean(dailyTotal$steps, na.rm=TRUE)
medianSteps <- median(dailyTotal$steps, na.rm=TRUE)
```

The mean is `r sprintf( "%.2f", meanSteps)`.
The median is `r sprintf( "%.2f", medianSteps)`.

## What is the average daily activity pattern?

1. Create a new table containing average steps by date
```{r}
dailyAverage <- ddply( data[ c(1,3)], .(interval), colwise(mean), na.rm=TRUE )
```

2. Plot a time series of the 5-minute interval against the average number of steps taken
```{r}
plot( dailyAverage$interval, dailyAverage$steps, type="l", main="Average Daily Pattern", xlab="5-minute time interval", ylab="Average steps")
```

3. Identify the 5-minute interval with the highest average number of steps.
```{r}
maxStepsInterval <- dailyAverage$interval[ which.max( dailyAverage$steps )]
maxSteps <- dailyAverage$steps[ which.max( dailyAverage$steps )]
```

The 5-minute interval with the highest number of steps is interval `r maxStepsInterval` with an average of `r sprintf( "%.2f", maxSteps)` steps.

## Imputing missing values

1. Determine the total number of rows with missing (NA) data.
```{r}
summary(data)
```

"Steps"" is the only column with missing data. Retrieve the number of rows.
```{r}
rowsWithNA <- summary( data$steps )[ c("NA's" ) ]
```
There are `r rowsWithNA[ "NA's" ]` rows with missing data.

2. Create a new dataset based on the original with the missing data filled with the daily average for the 5-minute interval
```{r}
test <- function(s,t) {
    if ( is.na(s) )
      return( dailyAverage$steps[ dailyAverage$interval == t ] )
    else
      return( s )
  }

newData <- ddply( data, .(steps,interval,date), transform, steps = test( steps, interval ),
                   interval=interval, date=date )
summary(newData)
```

3. Plot a histogram of the new dataset

```{r}
#Create a new table containing total steps by date
newDailyTotal <- ddply( newData, .(date), summarize, steps = sum( steps ) )

#Plot a histogram of the total number of steps taken each day
hist( newDailyTotal$steps, breaks=12, main="Histogram of total number of steps taken per day", xlab="Steps per Day" )

#Compute the mean and median total number of steps per day.
newMeanSteps <- mean(newDailyTotal$steps, na.rm=TRUE)
newMedianSteps <- median(newDailyTotal$steps, na.rm=TRUE)
```

The mean is `r sprintf( "%.2f", newMeanSteps)`.
The median is `r sprintf( "%.2f", newMedianSteps)`.

The mean does not differ from the original mean of `r sprintf( "%.2f", meanSteps)`.
There is a slight change from the original median of `r sprintf( "%.2f", medianSteps)`.

Imputing missing data has altered the data by increasing the frequency of occurence as evident in the new histogram.

## Are there differences in activity patterns between weekdays and weekends?

1. Introduce a two-level factor variable indicating "weekday" and "weekend"
```{r}
newData$dayType <- "weekday"
newData$dayType[ weekdays( newData$date ) %in% c( "Saturday", "Sunday" ) ] <- "weekend"
newData$dayType <- as.factor( newData$dayType )
```

2. Build a two-panel time series plot comparing "weekend" and "weekday" 5-minute interval against average number of steps.
```{r}
# Create a new table containing average steps by day type
dayTypeAverage <- ddply( newData[ c(1,3,4)], .(interval,dayType), colwise(mean), na.rm=TRUE )

# Build lattice plot
library( lattice )
xyplot( steps ~ interval | dayType,
        data = dayTypeAverage,
        type = "l",
        layout = c( 1, 2 ),
        main = "Weekend vs weekday comparison",
        ylab = "Number of steps",
        xlab = "5-minute interval"
        )
```